Staged Abstract Interpreters

### 2018-11-12 paper/21_bg_interp.tex
paper

\paragraph{Generic Interface.} 
Before showing the actual implementation of the concrete semantics, let's first see 
a generic interface for the interpreters, which will be used later for both unstaged/staged and concrete/abstract scenarios.
The trait \texttt{Semantics} first declares several abstract type members, such as \texttt{Addr},
\texttt{Value}, \texttt{Env}, \texttt{Store} and returned type \texttt{Ans}. 
Initial values for environments and stores, and few abstract methods that manipulate environments and stores are also declared.

>>>>

\paragraph{Generic Interface.} 
Before showing the actual implementation of the concrete semantics, let's first see 
a generic interface for the interpreters, which is semantics-agnostic and can be 
"interpreted" by either concrete or abstract, unstaged or staged semantics by our choice.
The trait \texttt{Semantics} first declares several abstract type members, such as \texttt{Addr},
\texttt{Value}, \texttt{Env}, \texttt{Store} and returned type \texttt{Ans}. 
Initial values for environments and stores, and few abstract methods that manipulate environments and stores are also declared.

### 2018-11-12 paper/23_bg_stage.tex
paper

\paragraph{Staged Concrete Semantics.} In the staging part, we share the same concrete type instantiation, 
i.e. trait \texttt{Concrete}, but we reimplement the staged version for the concrete operations.
Meanwhile, @RepConcInterpOps@ also extends from @LMSOps@ which provides necessary supports for the 
staged version of primitive types and data structures.

\begin{lstlisting}
trait RepConcInterpOps extends Concrete with LMSOps {
  type R[+T] = Rep[T]
  val $\rho$0: Rep[Env] = Map[Ident,Addr]()
  val $\sigma$0: Rep[Store] = Map[Addr,Value]()
  def get($\rho$: Rep[Env], x: Ident): Rep[Addr] = $\rho$(x)
  def put($\rho$: Rep[Env], x: Ident, a: Rep[Addr]): Rep[Env] = 
    $\rho$ + (unit(x) -> a)
  def get($\sigma$: Rep[Store], a: Rep[Addr]): Rep[Value] = $\sigma$(a)
  def put($\sigma$: Rep[Store], a: Rep[Addr], v: Rep[Value]): Rep[Store] = 
    $\sigma$ + (a -> v)
  def alloc($\sigma$: Rep[Store], x: Ident): Rep[Addr] = $\sigma$.size + 1
  def num(i: Lit): Rep[Value] = NumV(i)
  def branch0(cnd: Rep[Value], thn: => Ans, els: => Ans): Ans = {
    //FIXME: the cast is ugly, patten match?
    val i = cnd.asInstanceOf[Rep[NumV]].i
    if (i == 0) thn else els
  }
  def prim_eval(op: Symbol, 
    v1: Rep[Value], v2: Rep[Value]): Rep[Value] = op match {
      case '+ => v1.asInstanceOf[Rep[NumV]].i + 
                 v2.asInstanceOf[Rep[NumV]].i
      ...
    }
  ...
}
\end{lstlisting}

The first notable change is the abstract type member @R@ is assigned to be @Rep@, and accordingly the affected 
types such as @Env@ and @Store@ become @Rep[Env]@ and @Rep[Store]@. \todo{NumV}
The interesting thing is how we handle closures and function applications when staging is involved.
In the unstaged version, function @close@ is used to lift a literal lambda term to a closure, 
however, what we desire is the a \textit{compiled} closure, i.e., the syntactic term of the body expression should be
eliminated and specialized away.
The specialization of the interpreter with respect to the body of the lambda term proceed under the
assumptions that the argument and latest store will be provided later. \note{to denotational style?, back to HO?}
We will see this is an important observation that enables us to specialize an abstract interpreter modularly.
In the end of @close@, we create a @CompiledClo@ object that contains a staged function, which both will be
exhibited in the generated code.

>>>>

\paragraph{Staged Concrete Semantics.} In the staging part, we share the same concrete type instantiation, 
i.e. trait \texttt{Concrete}, but we reimplement the staged version for the concrete operations.
Meanwhile, @RepConcInterpOps@ also extends from @LMSOps@ which provides necessary supports for the 
staged version of primitive types and data structures. \todo{what are static, what are dyanmic?}

\begin{lstlisting}
trait RepConcInterpOps extends Concrete with LMSOps {
  type R[+T] = Rep[T]
  val $\rho$0: Rep[Env] = Map[Ident,Addr]()
  val $\sigma$0: Rep[Store] = Map[Addr,Value]()
  def get($\rho$: Rep[Env], x: Ident): Rep[Addr] = $\rho$(x)
  def put($\rho$: Rep[Env], x: Ident, a: Rep[Addr]): Rep[Env] = 
    $\rho$ + (unit(x) -> a)
  def get($\sigma$: Rep[Store], a: Rep[Addr]): Rep[Value] = $\sigma$(a)
  def put($\sigma$: Rep[Store], a: Rep[Addr], v: Rep[Value]): Rep[Store] = 
    $\sigma$ + (a -> v)
  def alloc($\sigma$: Rep[Store], x: Ident): Rep[Addr] = $\sigma$.size + 1
  def num(i: Lit): Rep[Value] = lift(NumV(i))
  def branch0(cnd: Rep[Value], thn: => Ans, els: => Ans): Ans = {
    //FIXME: the cast is ugly, patten match?
    val i = cnd.asInstanceOf[Rep[NumV]].i
    if (i == 0) thn else els
  }
  def prim_eval(op: Symbol, 
    v1: Rep[Value], v2: Rep[Value]): Rep[Value] = op match {
      case '+ => v1.asInstanceOf[Rep[NumV]].i + 
                 v2.asInstanceOf[Rep[NumV]].i
      ...
    }
  ...
}
\end{lstlisting}

The first notable change is the abstract type member @R@ is assigned to be @Rep@, and accordingly the affected 
types such as @Env@ and @Store@ become @Rep[Env]@ and @Rep[Store]@. For @num@, we use @lift@ function which is a
built-in in LMS that lifts a current-stage value to next stage.
The interesting thing is how we handle closures and function applications when staging is involved.
In the unstaged version, function @close@ is used to lift a literal lambda term to a closure, which
is a pair of the syntactic lambda term and the enclosing environment.
However, what we desire is the a \textit{compiled} closure, instead of a lifted @CloV@ value like 
what we did for @NumV@ --- in other words, the syntactic term of the lambda expression should be 
eliminated and specialized away.
The specialization of the interpreter with respect to the body of the lambda term proceed under the
assumptions that the argument and latest store will be provided later. \note{to denotational style?, back to HO?}
We will see this is an important observation that enables us to specialize an abstract interpreter modularly.
In the end of @close@, we create a @CompiledClo@ object that contains a staged function, which both will be
exhibited in the generated code.

### 2018-11-12 paper/30_sai.tex
paper

\paragraph{Staged Lattices.} The lattice definition we introduced in Section~\ref{unstaged_abs}
leaves space to be stage polymorphic, now we still use powerset as an example to present its 
staged version.

@@

\paragraph{Staged Abstract Semantics.}

\begin{lstlisting}
trait RepAbsInterpOps extends Abstract with LMSOps {
  type R[+T] = Rep[T]
  val $\rho$0: Rep[Env] = Map[Ident, Addr]()
  val $\sigma$0: Rep[Store] = Map[Addr, Value]()
  def get($\rho$: Rep[Env], x: Ident): Rep[Addr] = $\rho$(x)
  def put($\rho$: Rep[Env], x: Ident, a: Rep[Addr]): 
    Rep[Env] = $\rho$ + (unit(x) -> a)
  def get($\sigma$: Rep[Store], a: Rep[Addr]): Rep[Value] = 
    $\sigma$.getOrElse(a, RepLattice[Value].bot)
  def put($\sigma$: Rep[Store], a: Rep[Addr], v: Rep[Value]): 
    Rep[Store] = $\sigma$ + (a -> RepLattice[Value].$\sqcup$(v, get($\sigma$, a)))
  def alloc($\sigma$: Rep[Store], x: Ident): Rep[Addr] = Addr(x)
  def num(i: Lit): Rep[Value] = Set(NumV())
  def close(ev: EvalFun)($\lambda$: Lam, $\rho$: Rep[Env]): Rep[Value] = {
    val Lam(x, e) = $\lambda$
    val f: Rep[(Value, Store)]=>Rep[(Value,Store)] = 
      (as: Rep[(Value,Store)]) => {
        val args = as._1; val $\sigma$ = as._2; val $\alpha$ = alloc($\sigma$, x)
        ev(e, put($\rho$, x, $\alpha$), put($\sigma$, $\alpha$, args))
      }
    Set[AbsValue](CompiledClo(fun(f)))
  }
  def branch0(cnd: Rep[Value], thn: => Ans, els: => Ans): Ans =
    thn $\sqcup$ els
  def prim_eval(op: Symbol, 
                v1: Rep[Value], v2: Rep[Value]): Rep[Value] = 
    Set(NumV())
  ...
}
\end{lstlisting}

\paragraph{Staged Fixpoint Iteration.} staged fixed point iteration with memoization

\paragraph{Specialized Data Structures.}

>>>>

\paragraph{Staged Lattices.} In Section~\ref{unstaged_abs}, we exploited the higher-kinded type @R@
to leave space for staging lattices, now we instantiate @R@ to @Rep@ and still use powerset as an 
example to present its staged version.

@@

\paragraph{Staged Abstract Semantics.} We have seen how to obtain a staged concrete semantics
based on types, now we take the same approach to obtain a staged abstract semantics.
\todo{what are static, what are dynamic?}
The basic operations largely kept same as the unstaged version, except the types are changed to @Rep@.
Besides, when we update the environment, the identifier @x@ is known statically, but the map of environment
has type @Rep[Map[Ident,Addr]]@, so we apply @lift@ to @x@ to make it as a next-stage value.

\begin{lstlisting}
trait RepAbsInterpOps extends Abstract with LMSOps {
  type R[+T] = Rep[T]
  val $\rho$0: Rep[Env] = Map[Ident, Addr]()
  val $\sigma$0: Rep[Store] = Map[Addr, Value]()
  def get($\rho$: Rep[Env], x: Ident): Rep[Addr] = $\rho$(x)
  def put($\rho$: Rep[Env], x: Ident, a: Rep[Addr]): 
    Rep[Env] = $\rho$ + (lift(x) -> a)
  def get($\sigma$: Rep[Store], a: Rep[Addr]): Rep[Value] = 
    $\sigma$.getOrElse(a, RepLattice[Value].bot)
  def put($\sigma$: Rep[Store], a: Rep[Addr], v: Rep[Value]): 
    Rep[Store] = $\sigma$ + (a -> RepLattice[Value].$\sqcup$(v, get($\sigma$, a)))
  def alloc($\sigma$: Rep[Store], x: Ident): Rep[Addr] = Addr(x)
  def num(i: Lit): Rep[Value] = Set(NumV())
  def branch0(cnd: Rep[Value], thn: => Ans, els: => Ans): Ans =
    thn $\sqcup$ els
  def prim_eval(op: Symbol, 
                v1: Rep[Value], v2: Rep[Value]): Rep[Value] = 
    Set(NumV())
  ...
}
\end{lstlisting}

Once more, the way we handle closures is as same as the staged concrete interpreter:
the recursive call to @ev@ with the body expression @e@ is compiled and specialized, 
the wrapper function @f@ and will be a field value of @CompiledClo@ and be generated
for the next stage. At the end, we return a singleton set.

\begin{lstlisting}
def close(ev: EvalFun)($\lambda$: Lam, $\rho$: Rep[Env]): Rep[Value] = {
  val Lam(x, e) = $\lambda$
  val f: Rep[(Value, Store)]=>Rep[(Value,Store)] = {
    case (args: Rep[Value], $\sigma$: Rep[Store]) =>
      val args = as._1; val $\sigma$ = as._2; val $\alpha$ = alloc($\sigma$, x)
      ev(e, put($\rho$, x, $\alpha$), put($\sigma$, $\alpha$, args))
    }
  Set[AbsValue](CompiledClo(fun(f)))
}
def apply_closure(ev: EvalFun)
  (f: Rep[Value], arg: Rep[Value], $\sigma$: Rep[Store]): Ans = {
    reflectEffect(ApplyClosure(f, arg, $\sigma$))
  }
\end{lstlisting}

When generating code for application, we can not directly apply the callee.
Instead, we emit code that calls a next-stage function @apply_closures_norep@ 
which non-deterministically applies multiple target closure with the argument
and latest store, and finally returns the joined value and a single store.

\begin{lstlisting}
case ApplyClosures(fs, arg, $\sigma$) =>
  emitValDef(sym, "apply_closures_norep(" + 
                  quote(fs) + "," + quote(arg) + 
                  "," + quote($\sigma$) + ")")
\end{lstlisting}

\paragraph{Staged Fixpoint Iteration.} 
\todo{staged fixed point iteration with memoization}

\subsection{Specialized Data Structures}

Now we already obtained a working version of staged abstract interpter that is able to 
specialize an analysis. However, we treat the data structures such as @Map@s as black-boxes,
which means any operations on a @Map@ becomes code in the next stage.
But, as we identified before \todo{where}, the keys of environment are identifiers in the program,
which are completely known statically. This leaves us a chance to further specialize the 
data structures. Particularly, if we are specializing a monovariant analysis, the address space 
is equivalent to the identifiers, then the environment component can be entirely eliminated, 
and the store is a partially specialized map.
\todo{maps to arrays}

### 2018-11-12 paper/50_cases.tex
section 2/3
paper

@@

\subsection{Control-flow Analysis}

@@

@@

\subsubsection{Store-widening}

>>>>

In Section~\ref{sai}, we have shown staging an abstract interpreter is feasible and can be 
systematic, despite the fact that the abstract interpreter is intended to be imprecise and easy to implement. 
In this section, we conduct several case studies to show that it is also useful and widely
applicable to different analysis.

@@

\subsection{Control-flow Analysis} \label{cfa}

@@

\subsubsection{Regain Precise Store}

@@
